---
title: "Title"
author: "Angiely Camacho, Jefferson Pastuña"
date: "2024-04-17"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

# Introduction

Introduction...

# Before to start

Before to start...

# Notame workflow

As a first step...

```{r echo=TRUE, message=FALSE}

# Notame package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("antonvsdata/notame", ref = "v0.3.1")

# Notame library call
library(notame)

# Dependency packages installation
install_dependencies

```

Then, a main path and a log system was added to have a record of each process executed.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Main path
ppath <- "../Amazonia_Honey/"
# Log system
init_log(log_file = paste0(ppath, "Result/notame_Result/HS_GCMS/HS_GCMS_log.txt"))

```

Next...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

data <- read_from_excel(file = "Data/Data_to_notame/HS_GCMS_Data_to_notame.xlsx",
                        sheet = 2, corner_row = 16, corner_column = "L",
                        split_by = c("Column", "Ion Mode"))

```

Once the data is read, the next step was...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

modes <- construct_metabosets(exprs = data$exprs, 
                              pheno_data = data$pheno_data, 
                              feature_data = data$feature_data,
                              group_col = "Group")

```

## Preprocessing

The first step is...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Data extraction
mode <- modes$RTX5MS_EI
# Change 0 value to NA
mode <- mark_nas(mode, value = 0)

```

Then...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Low detection rate
mode <- flag_detection(mode, qc_limit = 7/9, group_limit = 2/3)

```

The next step...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drift correction
corrected <- correct_drift(mode)
#corrected <- correct_drift(corrected)
# Flag low quality features
#corrected <- flag_quality(corrected, condition = "RSD_r < 0.3 & D_ratio_r < 0.6")

```

Then we can visualize the data after drift correction.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
corr_bp <- plot_sample_boxplots(corrected,
                                order_by = "Group",
                                fill_by = "Species")
# PCA
corr_pca <- plot_pca(corrected,
                     center = TRUE,
                     shape = "Species",
                     color = "Species")
# Package to plots visualization in a same windows
library(patchwork)
# Plot
corr_pca + corr_bp

```

The next step...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

clustered <- cluster_features(corrected, rt_window = 1/60, corr_thresh = 0.95,
                              d_thresh = 0.80)
compressed <- compress_clusters(clustered)

```

We can visualize data...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
clust_bp <- plot_sample_boxplots(compressed,
                                 order_by = "Group",
                                 fill_by = "Species")
# PCA
clust_pca <- plot_pca(compressed,
                      center = TRUE,
                      shape = "Species",
                      color = "Species")
# Plot
clust_pca + clust_bp

```

Finally the data is...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

#save(compressed, file = paste0(ppath, "Result/notame_Result/HS_GCMS/Notame_HS_GC-MS_out.RData"))

```

##Processing

Here read this paper: https://doi.org/10.1007%2Fs11306-016-1030-9

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Impute missing values using random forest
# To clean data
#set.seed(615)
#imputed <- impute_rf(compressed)
# To all data
#imputed <- impute_rf(imputed, all_features = TRUE)
# Impute missing values using noise threshold
imputed <- impute_simple(compressed, value = 45, na_limit = 0)

```

We can inspect PCA plot after...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
imp_bp <- plot_sample_boxplots(imputed,
                               order_by = "Group",
                               fill_by = "Species")
# PCA
imp_pca <- plot_pca(imputed,
                    center = TRUE,
                    shape = "Species",
                    color = "Species")
# Plot
imp_pca + imp_bp

```

The next step...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Probabilistic quotient normalization
pqn_set <- pqn_normalization(imputed,
                             ref = c("qc", "all"),
                             method = c("median", "mean"),
                             all_features = FALSE)

```

We can inspect PCA plot after...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
pqn_bp <- plot_sample_boxplots(pqn_set,
                               order_by = "Group",
                               fill_by = "Species")
# PCA
pqn_pca <- plot_pca(pqn_set,
                    center = TRUE,
                    shape = "Species",
                    color = "Species")
# Plot
pqn_pca + pqn_bp

```

# Principal component analysis (PCA)

Droping flagged features...

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Extract clean data
pqn_noflag <- drop_flagged(pqn_set)
# Extracting feature height table
peak_height <- exprs(pqn_noflag)
# Extracting phenotypic data
pheno_data <- pqn_noflag@phenoData@data

```

Preparing data and transposing feature table.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Transposing feature height table
transp_table  <- t(peak_height)
# Centering and Scaling features
ei_pca <- prcomp(transp_table, center = TRUE, scale. = TRUE)

```

Plotting PCA results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 7, fig.height = 4}

# Library to left_join use
library(dplyr)
# PCA scores
scores <- ei_pca$x %>%                   # Get PC coordinates
  data.frame %>%                         # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%    # Create a new column with the sample names
  left_join(pheno_data)                  # Adding metadata
# PCA plot
pca_plot <- ggplot(scores,
       aes(PC1, PC2, shape = Species, color = Species)) +
  geom_point(size = 3) +
  guides(x=guide_axis(title = "PC1 (20.14 %)"),
         y=guide_axis(title = "PC2 (17.62 %)")) +
  labs(shape = 'Bees species', color= 'Bees species') +
  theme_classic() +
  theme(legend.text = element_text(face="italic")) +
  theme(legend.position = c(0.120, 0.230),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
pca_plot

```

Plotting loading results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

loadings <- ei_pca$rotation %>%           # Extract loadings
  data.frame(Feature_ID = rownames(.))    # New column with feat name

```

Creating an artificial table with Feature name and Compound column.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 14, fig.height = 7}

# Extracting feature identified
metab_data <- pqn_noflag[!is.na(pqn_noflag@featureData@data$Metabolite),]
# Extracting metabolite table
meta_table <- metab_data@featureData@data
# Creating a new small table of the annotated compounds
ei_compouds <- left_join(meta_table, loadings)
# Plotting results
load_pca <- ggplot(loadings, aes(PC1, PC2)) + 
  geom_point(alpha = 0.3, size = 2) +
  theme_classic() + 
  geom_point(data = ei_compouds,
             aes(shape = meta_table$IL,
                 color = meta_table$IL),
             size = 2.5) +
  labs(shape = 'Identification level',
       color = 'Identification level') +
  ggrepel::geom_label_repel(data = ei_compouds,
                            aes(label = meta_table$Metabolite),
                            box.padding = 0.37,
                            label.padding = 0.22,
                            label.r = 0.30,
                            cex = 2.5,
                            max.overlaps = 50,
                            min.segment.length = 0) +
  guides(x=guide_axis(title = "PC1 (20.14 %)"),
         y=guide_axis(title = "PC2 (17.62 %)")) +
  theme(legend.position = c(0.070, 0.090),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray") +
  ggsci::scale_color_aaas()
load_pca
# Save plot
#ggsave('Result/notame_Result/HS_GCMS/load_pca.pdf', width = 14, height = 7, device='pdf', dpi="print")

```

# Heat map plot with HCA

ComplexHeatmap package and dependency installation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# ComplexHeatmap package installation
#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
#BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)

# ColorRamp2 package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("jokergoo/colorRamp2")
library(colorRamp2)

# Cowplot package installation
#install.packages("cowplot")
library(cowplot)

# mdatools package installation
#install_github('svkucheryavski/mdatools')
library(mdatools)

# ClassyFire package installation
#remotes::install_github('aberHRML/classyfireR')
library(classyfireR)

```

The metabolites were classified using the ClassyFireR to add these metabolite classifications to the heatmap plot.


```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# InChI key of the metabolites you want to classify
InChI_Keys <- c('2-Heptanone' = "CATSNJVOTSVZJV-UHFFFAOYSA-N")
# Get classification
Classification_List <- purrr::map(InChI_Keys, get_classification)
Classification_List

```

Extracting and loaded of identified metabolites abundance.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drop QC
hm_no_qc <- drop_qcs(pqn_noflag)
# Scaling by autoscaling method
hm_scl <- scale(t(exprs(hm_no_qc)), center = TRUE, scale = TRUE)
hm_scl <- t(hm_scl)
# Adding autoscaling data to notame MetaboSet
hm_scl_set <- hm_no_qc
exprs(hm_scl_set) <- hm_scl
# Extracting identified metabolite data
raw_hm <- hm_scl_set[!is.na(hm_scl_set@featureData@data$Metabolite),]
# Extracting feature height table
hm_height <- exprs(raw_hm)
# Extracting sample information
hm_pdata <- raw_hm@phenoData@data
# Extracting feature information
hm_fdata <- raw_hm@featureData@data

```

Scaling, row and top heatmap anotation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.width = 10, fig.height = 9}

set.seed(1540)
# Adding row and column names
hm_scl <- hm_height
rownames(hm_scl) <- hm_fdata$Metabolite
colnames(hm_scl) <- hm_pdata$Species
# Metabolite class color
cols_metclass <- c("Benzenoids" = "#800000FF",
                   "Hydrocarbons" = "#FFA319FF",
                   "Lipids and lipid-like molecules" = "#8A9045FF",
                   "Organic oxygen compounds" = "#8DD3C7",
                   "Organohalogen compounds" = "#BEBADA",
                   "Organoheterocyclic compounds" = "#FFFFB3")
# Add row anotation to HeatMap
hm_row_ann <- rowAnnotation(`Superclass` = hm_fdata$classyfireR_Superclass,
                            col = list(`Superclass` = cols_metclass),
                            show_annotation_name = T,
                            show_legend = F)
# Species color
cols_species <- c("T. angustula" = "#E76BF3",
                  "M. fasciculata" = "#F8766D",
                  "M. fuscopilosa" = "#7CAE00")
# Add top anotation to Heatmap
top_info_ann <- HeatmapAnnotation(`Species` = hm_pdata$Species,
                                  col = list(`Species` = cols_species),
                                  show_annotation_name = T,
                                  show_legend = F,
                                  border = T)
# Color scale
mycol <- colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))
# Heatmap matrix plotting
hm_plot <- Heatmap(hm_scl,
                   col = mycol,
                   border_gp = grid::gpar(col = "black", lty = 0.02),
                   rect_gp = grid::gpar(col = "black", lwd = 0.75),
                   clustering_distance_columns = "euclidean",
                   clustering_method_columns = "complete",
                   top_annotation = top_info_ann,
                   column_names_gp = gpar(fontface = "italic"),
                   row_names_max_width = unit(10, "cm"),
                   right_annotation = hm_row_ann,
                   show_heatmap_legend = F,
                   row_km = 3, column_km = 2,
                   row_title = c("a", "b", "c"))
hm_plot

```

Adding legends to heatmap.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Color scale legend
lgd1 <- Legend(col_fun = mycol,
               title = "Autoscaled abundance",
               direction = "horizontal" )
# Bees species legend
lgd2 <- Legend(labels = gt_render(c("*T. angustula*",
                                    "*M. fasciculata*",
                                    "*M. fuscopilosa*")),
               legend_gp = gpar(fill = cols_species),
               title = "Bees species", ncol = 1)
# Metabolite class Legend
lgd3 <- Legend(labels = c(unique(hm_fdata$classyfireR_Superclass)) ,
               legend_gp = gpar(fill = cols_metclass), 
               title = "Metabolite superclass", ncol = 2)

```

ComplexHeatmap plot

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.width = 10, fig.height = 12}

set.seed(1540)
# Converting to ggplot
gg_heatmap <- grid.grabExpr(draw(hm_plot))
gg_heatmap <- ggpubr::as_ggplot(gg_heatmap)
# Legends
all_legends <- packLegend(lgd1, lgd2, lgd3, direction = "horizontal")
gg_legend <- grid.grabExpr(draw(all_legends))
gg_legend_fn <- ggpubr::as_ggplot(gg_legend)
# Heatmap plot
gcms_hm <- plot_grid(gg_legend_fn,
                     gg_heatmap, ncol = 1,
                     rel_heights = c(0.055, 0.880))
gcms_hm

```

# Correlation with the antimicrobial activity

Correlation between feature abundance and antibacterial result (zone inhibition diameter) using the PCA and Pearson correlation approaches.

## Using the PCA

Data and metadata preparation, merging the antimicrobial data with feature abundance.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Antimicrobial activity data extraction
bacteria_data <- drop_qcs(modes$na_na)
bacteria_data <- t(exprs(bacteria_data))
# Drop QCs of metabolomics data
no_qc <- drop_qcs(pqn_noflag)
# Extracting feature height table
corr_peak <- t(exprs(no_qc))
# Merge the antimicrobial activity result with the feature list table
bacteria_peak <- cbind(bacteria_data, corr_peak)
# Extracting phenotypic data of metabolomics data
corr_pdata <- no_qc@phenoData@data

```

### Score PCA

PCA scores calculation of the antimicrobial and feature abundance merge data.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Centering and Scaling features
corr_pca <- prcomp(bacteria_peak, center = TRUE, scale. = TRUE)

```

Plotting PCA score.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 7, fig.height = 4}

# PCA scores
corr_scores <- corr_pca$x %>%            # Get PC coordinates
  data.frame %>%                         # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%    # Create a new column with the sample names
  left_join(corr_pdata)                  # Adding metadata
# PCA plot
corr_pca_plot <- ggplot(corr_scores,
                        aes(PC1, PC2, shape = Species, color = Species)) +
  scale_color_manual(values=c("#F8766D",
                              "#7CAE00",
                              "#E76BF3")) +
  scale_shape_manual(values=c(16, 17, 3)) +
  geom_point(size = 3) +
  guides(x=guide_axis(title = "PC1 (22.46 %)"),
         y=guide_axis(title = "PC2 (18.82 %)")) +
  labs(shape = 'Bees species', color= 'Bees species') +
  theme_classic() +
  theme(legend.text = element_text(face="italic")) +
  theme(legend.position = c(0.120, 0.200),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
corr_pca_plot

```

### Loading PCA

Plotting loading results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

corr_loadings <- corr_pca$rotation %>%    # Extract loadings
  data.frame(Feature_ID = rownames(.))    # New column with feat name

```

Creating an artificial table with feature name and compound column.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 14, fig.height = 7}

# Extracting the antimicrobial metadata
bacteria_table <- drop_qcs(modes$na_na)@featureData@data
bacteria_table$Flag  <- as.character(bacteria_table$Flag)
# Merge the antimicrobial metadata with the metabolite table
corr_table <- merge(meta_table, bacteria_table, all = TRUE)
# Creating a new small table of the annotated compounds and antimicrobial metadata
corr_compouds <- left_join(corr_table, corr_loadings)
# Plotting results
corr_load_plot <- ggplot(corr_loadings, aes(PC1, PC2)) + 
  geom_point(alpha = 0.3, size = 2) +
  theme_classic() + 
  geom_point(data = corr_compouds,
             aes(shape = corr_table$IL,
                 color = corr_table$IL),
             size = 2.5) +
  scale_shape_manual(values = c(15, 16, 17)) +
  labs(shape = 'Legend of features',
       color = 'Legend of features') +
  ggrepel::geom_label_repel(data = corr_compouds,
                            aes(label = corr_table$Metabolite),
                            box.padding = 0.37,
                            label.padding = 0.22,
                            label.r = 0.30,
                            cex = 2.5,
                            max.overlaps = 50,
                            min.segment.length = 0) +
  guides(x=guide_axis(title = "PC1 (22.46 %)"),
         y=guide_axis(title = "PC2 (18.82 %)")) +
  theme(legend.position = c(0.070, 0.107),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray") +
  ggsci::scale_color_aaas()
corr_load_plot

```

## Using the Pearson correlation

The Pearson correlation between antimicrobial activity and feature abundance was performed.

### Normality of the antimicrobial data

First, the antimicrobial data normality was evaluated using the Shapiro-Wilk test.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Loading the antimicrobial activity data
all_bacteria <- read_excel("Data/Antimicrobial_of_honey.xlsx", sheet = 1)
# Deleting the group column to perform Shapiro-Wilk test in batch
no_group <- all_bacteria[,-1]
# Perform the Shapiro-Wilk test
shapiro <- do.call(rbind, lapply(no_group,
                                 function(x) shapiro.test(x)[c("statistic", "p.value")]))
shapiro

```

The result showed that only the microorganisms *K. pneumonia* and *P. mirabilis* have a normal distribution (p-value > 0.05). However, when we inspected the data, we found that the values in the ATA001 and RIG003 group in the *E. coli* could be outliers; also, a possible outlier in the RYMG001 group in the *S. aureus* was found. When we delete all the possible outliers, we find that all the microorganisms (not including *S. aureus* at 25 % v/v) have a normal distribution, as seen below.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Delete outlier data in E. coli
no_outliers <- all_bacteria
no_outliers[no_outliers$Group == "ATA001",]$E_coli_75 <- NA
no_outliers[no_outliers$Group == "ATA001",]$E_coli_100 <- NA
no_outliers[no_outliers$Group == "RIG003",]$E_coli_75 <- NA
no_outliers[no_outliers$Group == "RIG003",]$E_coli_100 <- NA
# Delete outlier data in S. aureus
no_outliers[no_outliers$Group == "RYMG001",]$S_aureus_25 <- NA
no_outliers[no_outliers$Group == "RYMG001",]$S_aureus_50 <- NA
no_outliers[no_outliers$Group == "RYMG001",]$S_aureus_75 <- NA
no_outliers[no_outliers$Group == "RYMG001",]$S_aureus_100 <- NA
# Deleting the group column to perform Shapiro-Wilk test in batch
no_outliers <- no_outliers[,-1]
# Perform the Shapiro-Wilk test
shapiro_outl <- do.call(rbind, lapply(no_outliers,
                                 function(x) shapiro.test(x)[c("statistic", "p.value")]))
shapiro_outl

```

The antimicrobial activity result of honey at 75 % v/v was used in the further analysis because this concentration showed activity in most of the tested microorganisms. (*E. coli*, *K. pneumonia*, *P. mirabilis*, and *S. aureus*).

### Metabolomics data transformation

Before the Pearson correlation of feature abundance with the antimicrobial activity (zone inhibition diameter), the previously normalized data by probabilistic quotient normalization (PQN) was transformed using the generalised logarithm (glog) method.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# "SummarizedExperiment" package installation
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("SummarizedExperiment")
library(SummarizedExperiment)
# Convert feature height table to SummarizedExperiment class
pmp_data <- SummarizedExperiment(assays = exprs(pqn_noflag),
                                 colData = pqn_noflag@phenoData@data)
# Package for generalized logarithmic transform
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("pmp")
library(pmp)
# Generalised logarithmic transform
glog_exprs <- glog_transformation(df = pmp_data@assays@data@listData[[1]],
                                  classes = pmp_data$QC,
                                  qc_label = "QC")
# Adding glog transformation to notame MetaboSet
glog_set <- pqn_noflag
exprs(glog_set) <- glog_exprs

```

Preparing the metabolomics data for the antimicrobial correlation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drop QCs
glog_no_qc <- drop_qcs(glog_set)
# Preparing the data for E. coli correlation (drop outlier data)
glog_E.coli <- glog_no_qc[, glog_no_qc$Group != "ATA001"]
pData(glog_E.coli) <- droplevels(pData(glog_E.coli))
glog_E.coli <- glog_E.coli[, glog_E.coli$Group != "RIG003"]
pData(glog_E.coli) <- droplevels(pData(glog_E.coli))
# Preparing the data for S. aureus (dropn outlier data)
glog_S.aureus <- glog_no_qc[, glog_no_qc$Group != "RYMG001"]
pData(glog_S.aureus) <- droplevels(pData(glog_S.aureus))
# Convert antimicrobial result to numeric
glog_E.coli$E_coli_75 <- as.numeric(glog_E.coli$E_coli_75)
glog_no_qc$K_pneumonia_75 <- as.numeric(glog_no_qc$K_pneumonia_75)
glog_no_qc$P_mirabilis_75 <- as.numeric(glog_no_qc$P_mirabilis_75)
glog_S.aureus$S_aureus_75 <- as.numeric(glog_S.aureus$S_aureus_75)

```

Perform the Pearson correlation analysis.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Feature correlation with the E. coli antimicrobial activity
E.coli_corr <- perform_correlation_tests(glog_E.coli,
                                         x = featureNames(glog_no_qc),
                                         y = c("E_coli_75"),
                                         method = "pearson")
# Feature correlation with the K. pneumonia antimicrobial activity
K.pneumonia_corr <- perform_correlation_tests(glog_no_qc,
                                              x = featureNames(glog_set),
                                              y = c("K_pneumonia_75"),
                                              method = "pearson")
# Feature correlation with the P. mirabilis antimicrobial activity
P.mirabilis_corr <- perform_correlation_tests(glog_no_qc,
                                              x = featureNames(glog_set),
                                              y = c("P_mirabilis_75"),
                                              method = "pearson")
# Feature correlation with the S. aureus antimicrobial activity
S.aureus_corr <- perform_correlation_tests(glog_S.aureus,
                                           x = featureNames(glog_no_qc),
                                           y = c("S_aureus_75"),
                                           method = "pearson")

```

### Correlation of identified metabolites

Create a matrix for the heatmap of correlation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Correlation table for the heatmap
hm_corr_table <- data.frame(Feature_ID = E.coli_corr$X,
                            E_coli = E.coli_corr$Correlation_coefficient,
                            K_pneumonia = K.pneumonia_corr$Correlation_coefficient,
                            P_mirabilis = P.mirabilis_corr$Correlation_coefficient,
                            S_aureus = S.aureus_corr$Correlation_coefficient)
# Feature data table
feat_corr_table <- data.frame(Feature_ID = hm_fdata$Feature_ID,
                             Metabolite = hm_fdata$Metabolite)
# Adding the metabolie name to the correlation table
hm_corr_table <- left_join(hm_corr_table, feat_corr_table)
# Extracting data of the identified metabolites
hm_corr_table <- hm_corr_table[!is.na(hm_corr_table$Metabolite),]
# Adding row name
rownames(hm_corr_table) <- hm_corr_table$Metabolite
# Delete extra information
hm_corr_table <- subset(hm_corr_table, select = -c(Feature_ID, Metabolite) )
# Converting DataFrame to data matrix
hm_corr_table <- data.matrix(hm_corr_table, rownames.force = NA)
# Adding column name
colnames(hm_corr_table) <- c("E. coli", "K. pneumonia", "P. mirabilis", "S. aureus")

```

Create a matrix showing the statistical significance of the correlation result.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Correlation table for the heatmap
hm_sign_table <- data.frame(Feature_ID = E.coli_corr$X,
                            E_coli = E.coli_corr$Correlation_P,
                            K_pneumonia = K.pneumonia_corr$Correlation_P,
                            P_mirabilis = P.mirabilis_corr$Correlation_P,
                            S_aureus = S.aureus_corr$Correlation_P)
# Adding the metabolie name to the correlation table
hm_sign_table <- left_join(hm_sign_table, feat_corr_table)
# Extracting data of the identified metabolites
hm_sign_table <- hm_sign_table[!is.na(hm_sign_table$Metabolite),]
# Adding row name
rownames(hm_sign_table) <- hm_sign_table$Metabolite
# Delete extra information
hm_sign_table <- subset(hm_sign_table, select = -c(Feature_ID, Metabolite) )
# Converting DataFrame to data matrix
hm_sign_table <- data.matrix(hm_sign_table, rownames.force = NA)
# Adding column name
colnames(hm_sign_table) <- c("E. coli", "K. pneumonia", "P. mirabilis", "S. aureus")

```

Plot the correlation heatmap.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.width = 5, fig.height = 9}

# Add top anotation to Heatmap
ann_corr <- HeatmapAnnotation(
  foo = anno_block(gp = gpar(fill = 0, col = "white"),
                   labels = "Pearson correlation",
                   labels_gp = gpar(col = "black", fontsize = 12,
                                    fontface = "bold")))
# Color scale
col_fun <- colorRamp2(c(-1, 0, 1), c("blue", "white", "red"))
# Heatmap plot
hm_corr <- Heatmap(hm_corr_table, col = col_fun, cluster_rows = FALSE,
                   cluster_columns = FALSE, column_names_gp = gpar(fontface = "italic"),
                   right_annotation = hm_row_ann,
                   border_gp = grid::gpar(col = "black", lty = 0.02),
                   rect_gp = grid::gpar(col = "black", lwd = 0.75),
                   top_annotation = ann_corr,
                   cell_fun = function(j, i, x, y, w, h, fill) {
                     if(hm_sign_table[i, j] < 0.05) {
                       grid.text(sprintf("%.1f", hm_corr_table[i, j]), x, y,
                                 gp = gpar(fontsize = 12, fontface = "bold"))
                       } else if(hm_sign_table[i, j] > 0.05) {
                         grid.text(sprintf("%.1f", hm_corr_table[i, j]), x, y,
                                   gp = gpar(fontsize = 12))}},
                   show_heatmap_legend = F)
hm_corr

```

### Correlation of unknown metabolites

Filtering the metabolites with positive and high Pearson correlation coefficient (r >0.5).

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Library to change character values
library('stringr')
# Correlation table for the heatmap with the best features
best_feat <- data.frame(Feature_ID = E.coli_corr$X,
                        E_coli = E.coli_corr$Correlation_coefficient,
                        E_coli_p = E.coli_corr$Correlation_P,
                        K_pneumonia = K.pneumonia_corr$Correlation_coefficient,
                        K_pneumonia_p = K.pneumonia_corr$Correlation_P,
                        P_mirabilis = P.mirabilis_corr$Correlation_coefficient,
                        P_mirabilis_p = P.mirabilis_corr$Correlation_P,
                        S_aureus = S.aureus_corr$Correlation_coefficient,
                        S_aureus_p = S.aureus_corr$Correlation_P)
# Filtering the feature with the correlation >= 0.5
best_feat <- best_feat[best_feat$E_coli >= 0.5 |
                         best_feat$K_pneumonia >= 0.5 |
                         best_feat$P_mirabilis >= 0.5 |
                         best_feat$S_aureus >= 0.5, ]
# Adding the metabolite name to the correlation table
best_feat <- left_join(best_feat, feat_corr_table)
# Extracting data of no identified metabolites
best_feat <- best_feat[!complete.cases(best_feat$Metabolite),]
# Replace the column name and ionization type "RTX5MS_EI" with unknown
best_feat$Feature_ID <- sub('RTX5MS_EI_','Unknown-', best_feat$Feature_ID)
best_feat$Feature_ID <- str_replace_all(best_feat$Feature_ID, '_', '.')

```

Create a matrix for the heatmap of correlation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Correlation table for the heatmap
hm_corr_table1 <- data.frame(Feature_ID = best_feat$Feature_ID,
                             E_coli = best_feat$E_coli,
                             K_pneumonia = best_feat$K_pneumonia,
                             P_mirabilis = best_feat$P_mirabilis,
                             S_aureus = best_feat$S_aureus)
# Adding row name
rownames(hm_corr_table1) <- hm_corr_table1$Feature_ID
# Delete extra information
hm_corr_table1 <- subset(hm_corr_table1, select = -c(Feature_ID) )
# Converting DataFrame to data matrix
hm_corr_table1 <- data.matrix(hm_corr_table1, rownames.force = NA)
# Adding column name
colnames(hm_corr_table1) <- c("E. coli", "K. pneumonia", "P. mirabilis", "S. aureus")

```

Create a matrix showing the statistical significance of the correlation result.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Correlation table for the heatmap
hm_sign_table1 <- data.frame(Feature_ID = best_feat$Feature_ID,
                             E_coli = best_feat$E_coli_p,
                             K_pneumonia = best_feat$K_pneumonia_p,
                             P_mirabilis = best_feat$P_mirabilis_p,
                             S_aureus = best_feat$S_aureus_p)
# Adding row name
rownames(hm_sign_table1) <- hm_sign_table1$Feature_ID
# Delete extra information
hm_sign_table1 <- subset(hm_sign_table1, select = -c(Feature_ID) )
# Converting DataFrame to data matrix
hm_sign_table1 <- data.matrix(hm_sign_table1, rownames.force = NA)
# Adding column name
colnames(hm_sign_table1) <- c("E. coli", "K. pneumonia", "P. mirabilis", "S. aureus")

```

Plot the correlation heatmap.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.width = 5, fig.height = 6}

# Packed legends
pd = packLegend(lgd1a, direction = "horizontal")
# Heatmap plot
hm_corr1 <- Heatmap(hm_corr_table1, col = col_fun, cluster_rows = FALSE,
                   cluster_columns = FALSE, column_names_gp = gpar(fontface = "italic"),
                   border_gp = grid::gpar(col = "black", lty = 0.02),
                   rect_gp = grid::gpar(col = "black", lwd = 0.75),
                   top_annotation = ann_corr,
                   cell_fun = function(j, i, x, y, w, h, fill) {
                     if(hm_sign_table1[i, j] < 0.05) {
                       grid.text(sprintf("%.1f", hm_corr_table1[i, j]), x, y,
                                 gp = gpar(fontsize = 10, fontface = "bold"))
                       } else if(hm_sign_table1[i, j] > 0.05) {
                         grid.text(sprintf("%.1f", hm_corr_table1[i, j]), x, y,
                                   gp = gpar(fontsize = 10))}},
                   show_heatmap_legend = F)
hm_corr1
draw(pd, x = unit(7.5, "cm"), y = unit(1, "cm"), just = c("left", "bottom"))

```

Finish a record.

```{r}

finish_log()

```
