---
title: "Chemical characterization of volatile compounds and biological activity of stingless bee honeys from the Ecuadorian Amazon using Headspace Gas Chromatography-Mass Spectrometry - Statistics of antimicrobial activity"
author: "Jefferson Pastuña"
date: "2025-06-17"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

# Introduction

This R Script aims to record the procedure for analyzing the antimicrobial activity of honeys (at different concentrations, 25, 50, 75, and 100 % v/v) from three different species (*Melipona fasciculata*, *Melipona fuscopilosa*, and *Tetragonisca angustula*) against the following microbial strains: *Candida albicans* ATCC 10231, *Escherichia coli* ATCC 25922, *Staphylococcus aureus* ATCC 29213, and antibiotic-resistant *Klebsiella pneumoniae*, and *Proteus mirabilis*. The result of *C. albicans* was not used for statistical analysis because the antimicrobial activity was observed in only a few honey samples. The agar well diffusion method was used to measure the antimicrobial activity, and the diameter of the inhibition zone was used for statistical analysis.

# Data normality test

All necessary libraries were loaded.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Loading the "readxl" package
library(readxl)
# "agricolae" package installation and library loadding
#install.packages("agricolae", repos = "https://cran.r-project.org")
library(agricolae)
# Loading the "ggplot2" package
library(ggplot2)
# dunn.test library loadding
library(dunn.test)
# FSA library loadding
library(FSA)
# rcompanion library loadding
library(rcompanion)

```

First, the antimicrobial data normality was evaluated using the Shapiro-Wilk test.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Loading the antimicrobial activity data
all_bacteria <- read_excel("../Data/Antimicrobial_of_honey.xlsx", sheet = 2)
# Deleting the group column to perform Shapiro-Wilk test in batch
no_group <- all_bacteria[,-1]
# Perform the Shapiro-Wilk test
shapiro <- do.call(rbind, lapply(no_group,
                                 function(x) shapiro.test(x)[c("statistic", "p.value")]))
shapiro

```

# Data homoscedasticity test

Second, the antimicrobial data homoscedasticity was evaluated using the Bartlett test.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

cols_aa <- names(all_bacteria)[2:ncol(all_bacteria)]
# Perform the Bartlett test test
bartlett <- lapply(cols_aa,
                   function(x) bartlett.test(reformulate("Group", x), data = all_bacteria))
bartlett

```

The result showed that only the microorganism *P. mirabilis* at 75 and 100 % v/v have a normal distribution (Shapiro-Wilk test p-value > 0.05) and equality of variance (Bartlett test p-value > 0.05). Therefore, the parametric statistical test was performed on the *P. mirabilis* data, and nonparametric tests were performed on the rest of the data.

# Parametric statistical test

For statistical analysis of *P. mirabilis* at 75 and 100 % v/v, the analysis of variance ANOVA and Tukey's honest significant difference (HSD) test were used.

## *P. mirabilis*

### ANOVA test

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# ANOVA test of P. mirabilis at 75 % v/v
oneway.test(all_bacteria$P_mirabilis_75 ~ all_bacteria$Group,
            data = all_bacteria, var.equal = TRUE)
# ANOVA test of P. mirabilis at 100 % v/v
oneway.test(all_bacteria$P_mirabilis_100 ~ all_bacteria$Group,
            data = all_bacteria, var.equal = TRUE)

```

The ANOVA test showed that the antimicrobial activity of honey from three different bee species collected from three different colonies was significantly different.

### Tukey test

The Tukey test was performed to inspect which groups have significantly different means.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Tukey test of P. mirabilis at 75 % v/v
modelo_pm75 <- aov(all_bacteria$P_mirabilis_75 ~ all_bacteria$Group,
                  data = all_bacteria)
tukey_pm75 <- HSD.test(modelo_pm75, "all_bacteria$Group",
                       group = TRUE, console = TRUE)
# Plotting result
tk_pm75 <- ggplot(all_bacteria,
                  aes(all_bacteria$Group,
                      all_bacteria$P_mirabilis_75,
                      color = all_bacteria$Group)) +
  scale_color_manual(values=c("#A73030", "#8F7700", "#4A6990",
                              "#CD534C", "#EFC000", "#EFC000",
                              "#7AA6DC", "#A73030", "#8F7700")) +
  geom_boxplot() +
  stat_summary(geom = 'text',
               label = c("C","EF","A","CD","F","F","DE","B","BC"),
               fun.y = max, vjust = -0.5) +
  theme_classic() +
  labs(x = NULL, y = "Zone inhibition diameter (mm)") +
  theme(plot.title = element_text(face = "italic"),
        axis.text.x = element_text(angle = 50, hjust = 1)) +
  theme(legend.position="none")
tk_pm75

```

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Tukey test of of P. mirabilis at 100 % v/v
modelo_pm100 <- aov(all_bacteria$P_mirabilis_100 ~ all_bacteria$Group,
                    data = all_bacteria)
tukey_pm100 <- HSD.test(modelo_pm100, "all_bacteria$Group",
                        group = TRUE, console = TRUE)
# Plotting result
tk_pm100 <- ggplot(all_bacteria,
                   aes(all_bacteria$Group,
                       all_bacteria$P_mirabilis_100,
                       color = all_bacteria$Group)) +
  scale_color_manual(values=c("#A73030", "#8F7700", "#4A6990",
                              "#A73030", "#7AA6DC", "#CD534C", "#7AA6DC",
                              "#4A6990", "#4A6990")) +
  geom_boxplot() + ylim(10, 18) +
  stat_summary(geom = 'text',
               label = c("B","BC","A","B","D","CD","D","A","A"),
               fun.y = max, vjust = -0.5) +
  theme_classic() +
  labs(x = NULL, y = "Zone inhibition diameter (mm)") +
  theme(plot.title = element_text(face = "italic"),
        axis.text.x = element_text(angle = 60, hjust = 1)) +
  theme(legend.position="none")
tk_pm100

```

# Nonparametric statistical test

## *E. coli*

### Kruskal-Wallis test

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Kruskal-Wallis test of E. coli (Honey at 75 % v/v)
kruskal.test(all_bacteria$E_coli_75 ~ all_bacteria$Group, data = all_bacteria)
# Kruskal-Wallis test of E. coli (Honey at 100 % v/v)
kruskal.test(all_bacteria$E_coli_100 ~ all_bacteria$Group, data = all_bacteria)

```

The Kruskal-Wallis test showed that the *E. coli* antimicrobial activity of honey from three different bee species collected from three different colonies was significantly different.

### Dunn test

The Dunn test was performed to inspect which groups have significantly different means.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Dunn test of E. coli (Honey at 75 % v/v)
dunn.test(all_bacteria$E_coli_75, all_bacteria$Group, method = "bh", list = TRUE, table = FALSE)
# Compact letter display
dunn_ec75 = dunnTest(all_bacteria$E_coli_75 ~ all_bacteria$Group,
                     data = all_bacteria,
                     method = "bh")
dunn_ec75_pt = dunn_ec75$res
cldList(P.adj ~ Comparison,
        data = dunn_ec75_pt,
        threshold = 0.05)
# Dunn test of E. coli (Honey at 100 % v/v)
dunn.test(all_bacteria$E_coli_100, all_bacteria$Group, method = "bh", list = TRUE, table = FALSE)
# Compact letter display
dunn_ec100 = dunnTest(all_bacteria$E_coli_100 ~ all_bacteria$Group,
                     data = all_bacteria,
                     method = "bh")
dunn_ec100_pt = dunn_ec100$res
cldList(P.adj ~ Comparison,
        data = dunn_ec100_pt,
        threshold = 0.05)
# Plotting result
dn_ec100 <- ggplot(all_bacteria,
                   aes(all_bacteria$Group,
                       all_bacteria$E_coli_100,
                       color = all_bacteria$Group)) +
  scale_color_manual(values=c("#8F7700", "#8F7700", "#4A6990",
                              "#8F7700", "#A73030", "#CD534C", "#8F7700",
                              "#7AA6DC", "#8F7700")) +
  geom_boxplot() +
  stat_summary(geom = 'text',
               label = c("abc","abc","a","abc","b","bc","abc","ac","abc"),
               fun.y = max, vjust = -0.5) +
  theme_classic() +
  labs(x = NULL, y = "Zone inhibition diameter (mm)") +
  theme(plot.title = element_text(face = "italic"),
        axis.text.x = element_text(angle = 60, hjust = 1)) +
  theme(legend.position="none")
dn_ec100

```

## *S. aureus*

### Kruskal-Wallis test

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Kruskal-Wallis test of S. aureus (Honey at 25 % v/v)
kruskal.test(all_bacteria$S_aureus_25 ~ all_bacteria$Group, data = all_bacteria)
# Kruskal-Wallis test of S. aureus (Honey at 50 % v/v)
kruskal.test(all_bacteria$S_aureus_50 ~ all_bacteria$Group, data = all_bacteria)
# Kruskal-Wallis test of S. aureus (Honey at 75 % v/v)
kruskal.test(all_bacteria$S_aureus_75 ~ all_bacteria$Group, data = all_bacteria)
# Kruskal-Wallis test of S. aureus (Honey at 100 % v/v)
kruskal.test(all_bacteria$S_aureus_100 ~ all_bacteria$Group, data = all_bacteria)

```

The Kruskal-Wallis test showed that the *S. aureus* antimicrobial activity of honey from three different bee species collected from three different colonies was significantly different.

### Dunn test

The Dunn test was performed to inspect which groups have significantly different means.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Dunn test of S. aureus (Honey at 25 % v/v)
dunn.test(all_bacteria$S_aureus_25, all_bacteria$Group, method = "bh", list = TRUE, table = FALSE)
# Compact letter display
dunn_sa25 = dunnTest(all_bacteria$S_aureus_25 ~ all_bacteria$Group,
                     data = all_bacteria,
                     method = "bh")
dunn_sa25_pt = dunn_sa25$res
cldList(P.adj ~ Comparison,
        data = dunn_sa25_pt,
        threshold = 0.05)
# Dunn test of S. aureus (Honey at 50 % v/v)
dunn.test(all_bacteria$S_aureus_50, all_bacteria$Group, method = "bh", list = TRUE, table = FALSE)
# Compact letter display
dunn_sa50 = dunnTest(all_bacteria$S_aureus_50 ~ all_bacteria$Group,
                     data = all_bacteria,
                     method = "bh")
dunn_sa50_pt = dunn_sa50$res
cldList(P.adj ~ Comparison,
        data = dunn_sa50_pt,
        threshold = 0.05)

# Dunn test of S. aureus (Honey at 75 % v/v)
dunn.test(all_bacteria$S_aureus_75, all_bacteria$Group, method = "bh", list = TRUE, table = FALSE)
# Compact letter display
dunn_sa75 = dunnTest(all_bacteria$S_aureus_75 ~ all_bacteria$Group,
                     data = all_bacteria,
                     method = "bh")
dunn_sa75_pt = dunn_sa75$res
cldList(P.adj ~ Comparison,
        data = dunn_sa75_pt,
        threshold = 0.05)

# Dunn test of S. aureus (Honey at 100 % v/v)
dunn.test(all_bacteria$S_aureus_100, all_bacteria$Group, method = "bh", list = TRUE, table = FALSE)
# Compact letter display
dunn_sa100 = dunnTest(all_bacteria$S_aureus_100 ~ all_bacteria$Group,
                     data = all_bacteria,
                     method = "bh")
dunn_sa100_pt = dunn_sa100$res
cldList(P.adj ~ Comparison,
        data = dunn_sa100_pt,
        threshold = 0.05)
# Plotting result
dn_sa100 <- ggplot(all_bacteria,
                   aes(all_bacteria$Group,
                       all_bacteria$S_aureus_100,
                       color = all_bacteria$Group)) +
  scale_color_manual(values=c("#A73030", "#8F7700", "#4A6990",
                              "#8F7700", "#7AA6DC", "#CD534C", "#8F7700",
                              "#8F7700", "#A73030")) +
  geom_boxplot() +
  stat_summary(geom = 'text',
               label = c("ab","abc","a","abc","c","bc","abc","abc","ab"),
               fun.y = max, vjust = -0.5) +
  theme_classic() +
  labs(x = NULL, y = "Zone inhibition diameter (mm)") +
  theme(plot.title = element_text(face = "italic"),
        axis.text.x = element_text(angle = 60, hjust = 1)) +
  theme(legend.position="none")
dn_sa100

```

## *K. pneumoniae*

### Kruskal-Wallis test

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Kruskal-Wallis test of K. pneumoniae (Honey at 50 % v/v)
kruskal.test(all_bacteria$K_pneumonia_50 ~ all_bacteria$Group, data = all_bacteria)
# Kruskal-Wallis test of K. pneumoniae (Honey at 75 % v/v)
kruskal.test(all_bacteria$K_pneumonia_75 ~ all_bacteria$Group, data = all_bacteria)
# Kruskal-Wallis test of K. pneumoniae (Honey at 100 % v/v)
kruskal.test(all_bacteria$K_pneumonia_100 ~ all_bacteria$Group, data = all_bacteria)

```

The Kruskal-Wallis test showed that the *K. pneumoniae* antimicrobial activity of honey from three different bee species collected from three different colonies was significantly different.

### Dunn test

The Dunn test was performed to inspect which groups have significantly different means.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Dunn test of K. pneumoniae (Honey at 50 % v/v)
dunn.test(all_bacteria$K_pneumonia_50, all_bacteria$Group, method = "bh", list = TRUE, table = FALSE)
# Compact letter display
dunn_kp50 = dunnTest(all_bacteria$K_pneumonia_50 ~ all_bacteria$Group,
                     data = all_bacteria,
                     method = "bh")
dunn_kp50_pt = dunn_kp50$res
cldList(P.adj ~ Comparison,
        data = dunn_kp50_pt,
        threshold = 0.05)
# Dunn test of K. pneumoniae (Honey at 75 % v/v)
dunn.test(all_bacteria$K_pneumonia_75, all_bacteria$Group, method = "bh", list = TRUE, table = FALSE)
# Compact letter display
dunn_kp75 = dunnTest(all_bacteria$K_pneumonia_75 ~ all_bacteria$Group,
                     data = all_bacteria,
                     method = "bh")
dunn_kp75_pt = dunn_kp75$res
cldList(P.adj ~ Comparison,
        data = dunn_kp75_pt,
        threshold = 0.05)
# Dunn test of K. pneumoniae (Honey at 100 % v/v)
dunn.test(all_bacteria$K_pneumonia_100, all_bacteria$Group, method = "bh", list = TRUE, table = FALSE)
# Compact letter display
dunn_kp100 = dunnTest(all_bacteria$K_pneumonia_100 ~ all_bacteria$Group,
                     data = all_bacteria,
                     method = "bh")
dunn_kp100_pt = dunn_kp100$res
cldList(P.adj ~ Comparison,
        data = dunn_kp100_pt,
        threshold = 0.05)
# Plotting result
dn_kp100 <- ggplot(all_bacteria,
                   aes(all_bacteria$Group,
                       all_bacteria$K_pneumonia_100,
                       color = all_bacteria$Group)) +
  scale_color_manual(values=c("#4A6990", "#CD534C", "#CD534C",
                              "#CD534C", "#CD534C", "#CD534C", "#A73030",
                              "#CD534C", "#CD534C")) +
  geom_boxplot() + ylim(8, 16.2) +
  stat_summary(geom = 'text',
               label = c("a","ab","ab","ab","ab","ab","b","ab","ab"),
               fun.y = max, vjust = -0.5) +
  theme_classic() +
  labs(x = NULL, y = "Zone inhibition diameter (mm)") +
  theme(plot.title = element_text(face = "italic"),
        axis.text.x = element_text(angle = 60, hjust = 1)) +
  theme(legend.position="none")
dn_kp100

```

## *P. mirabilis*

### Kruskal-Wallis test

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Kruskal-Wallis test of P. mirabilis (Honey at 50 % v/v)
kruskal.test(all_bacteria$P_mirabilis_50 ~ all_bacteria$Group, data = all_bacteria)

```

The Kruskal-Wallis test showed that the *P. mirabilis* antimicrobial activity of honey from three different bee species collected from three different colonies was significantly different.

### Dunn test

The Dunn test was performed to inspect which groups have significantly different means.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Dunn test of P. mirabilis (Honey at 50 % v/v)
dunn.test(all_bacteria$P_mirabilis_50, all_bacteria$Group, method = "bh", list = TRUE, table = FALSE)
# Compact letter display
dunn_pm50 = dunnTest(all_bacteria$P_mirabilis_50 ~ all_bacteria$Group,
                     data = all_bacteria,
                     method = "bh")
dunn_pm50_pt = dunn_pm50$res
cldList(P.adj ~ Comparison,
        data = dunn_pm50_pt,
        threshold = 0.05)

```
