---
title: "Spectral deconvolution of stingless bee honeys HS-GC-MS metabolomics"
author: "Angiely Camacho, José Abata, Jefferson Pastuna"
date: "2024-04-12"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

# Introduction

The present document aims to record the HS-GC-EI (Q)MS spectral deconvolution procedure of the volatile profile of honeys of 3 native bee species (*Melipona fasciculata*, *Melipona fuscopilosa*, and *Tetragonisca angustula*) from 3 different localities in the Chontapunta parish, Tena, Ecuadorian Amazonia region.

The workflow used is taken from the paper [eRah: A Computational Tool Integrating Spectral Deconvolution and Alignment with Quantification and Identification of Metabolites in GC/MS-Based Metabolomics](https://doi.org/10.1021/acs.analchem.6b02927). It offers a wide variety of functions to automatically detect and deconvolve the spectra of the compounds appearing in GC–MS chromatograms. The "*eRah*" R package includes automatic peak detection, peak alignment, feature extraction, and compound identification.

# Before to start

The "*eRah*" R package accepts raw data files (netCDF or mzXML) obtained in GC–q/MS, GC-TOF/MS, and GC-qTOF/MS (using nominal mass) equipment.

In this case, the data was obtained with a GC-2030 gas chromatograph coupled to a GCMS-QP2020 NX quadrupole mass spectrometer operating in electron ionization (GC-EI-q/MS). The raw MS data (.qgd) were converted to (netCDF) format using the proprietary software of the instrument GCMS Postrun Analysis 4.53SP1. The data was organized in an experiment folder named Data_to_eRah. This contains three class-folders called ’Blank’, ’Quality_Control’, and ’Samples’, each containing the sample files for that class.

# eRah package workflow

The "*eRah*" R package workflow is based on five steps: (i) data pre-processing, (ii) spectral deconvolution, (iii) spectral alignment, (iv) retrieval of missing compounds, and (v) compound identification.

To use the "*eRah*" R package, you need to install the "*eRah*" package and call the library() function.

```{r echo=TRUE, message=FALSE, warning=FALSE}

# eRah package installation
#install.packages('erah')
# eRah library call
library(erah)

```

To proceed, you must delete the unwanted files in a specific path and create a directory with the wanted files.

```{r echo=TRUE}

# Delete all file that are not in folders
unlink('Data/Data_to_eRah/*')
# Data folder path
createdt('Data/Data_to_eRah/')

```

Load and process the necessary data from the CSV files, which contain the relevant information to proceed with the study about the volatilome of honey from different Amazonian stingless bees. The MetaboSet object was created to store the sample metadata and process.

**NOTE.** The instrumental and phenotype files created by eRah were relocated to a new directory. In both files, the semicolon delimiter has been changed to a comma delimiter. The complete directory path of the chromatograms has been added to the instrumental file.

```{r echo=TRUE}

# Loading (*.mzXML) chromatograms name (instrumental file)
#instrumental <- read.csv('Data/Metadata_to_eRah/Metadata_inst_mzXML.csv')
# If (*.mzXML) did not work
# Loading (*.CDF) chromatograms
instrumental <- read.csv('Data/Metadata_to_eRah/HS_GCMS_Data_inst_CDF.csv')
# Loading metadata of the chromatograms (phenotype file)
phenotype <- read.csv('Data/Metadata_to_eRah/HS_GCMS_Data_pheno.csv')
# Merge of metadata information with chromatograms
ex <- newExp(instrumental = instrumental,
             phenotype = phenotype,
             info = 'Amazonia honey')

```

## Compound deconvolution

It is a mathematical and computational technique used to recover an original signal from a complex pool of information.

Set up the parameters with the minimum height and width of the peak, the noise threshold, and the mz signals compounds wanted to exclude.

```{r echo=TRUE}

ex.dec.par <- setDecPar(min.peak.width = 2,
                        min.peak.height = 450,
                        noise.threshold = 45,
                        avoid.processing.mz = c(30:69,73:75,147:149))

```

To run parallel processing, the package "future" must be used.

**NOTE.** Adjusts the number of "workers" with the equipped PC cores.

```{r echo=TRUE}

plan(future::multisession,
     workers = 16)

```

We proceed to the deconvolution of compounds using the parameters specified in “dec_par”. The results obtained will be saved in (dec_data).

```{r echo=TRUE, warning=FALSE}

ex <- deconvolveComp(ex,
                     ex.dec.par)

```

## Alignment

Correcting for retention time variation of eluting compounds, facilitating relative quantification and comparison of compounds between samples, is done by grouping compounds within these limits of retention time distance and spectral similarity.

```{r echo=TRUE}

# Alignment parameters
ex.al.par <- setAlPar(min.spectra.cor = 0.80,
                      max.time.dist = 3,
                      mz.range = 70:550)
# Alignment
ex <- alignComp(ex,
                alParameters = ex.al.par)

```

## Missing compound recovery

Missing Compound Recovery is a function that aims to identify and recover information about compounds that may have been missed or undetected in the initial analysis due to various reasons, such as background noise, low signal intensity, or overlapping peaks.

To recover missing compounds, use the function "recMissComp" and save the result in "ex."

```{r echo=TRUE}

ex <- recMissComp(ex,
                  min.samples = 3)

```

Exporting alignment feature list.

```{r echo=TRUE}

# Extracting alignment feature list
feat_list <- alignList(ex,
                       by.area = FALSE)
# Exporting alignment feature list
#write.csv(feat_list,
#          file = "Result/eRah_Result/erah_Export_2Notame.csv")

```

# Identification

Metabolite identification was by comparing all the spectra found against a reference database. The "*eRah*" R package default database was used for metabolite identification. Metabolite identification was improved by exporting all the spectra found (.msp) in the "*eRah*" R package to the NIST MS Search 2.4 software.

```{r echo=TRUE, warning=FALSE}

# Identification
peak_iden <- identifyComp(ex,
                          id.database = mslib,
                          mz.range = NULL,
                          n.putative = 1)

```

Exporting spectra to NIST MS Search software for identification with the NIST-20 library.

```{r echo=TRUE}

#export2MSP(peak_iden,
#           store.path = "Result/eRah_Result",
#           alg.version = 2)

```

