---
title: "Chemical characterization of volatile compounds and biological activity of stingless bee honeys from the Ecuadorian Amazon using Headspace Gas Chromatography-Mass Spectrometry - Statistical analysis"
author: "Jefferson Pastuña"
date: "2024-04-17"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

# Introduction

This R Script aims to record the procedure for analyzing the volatile profile of honeys from three different stingless bee species (*Melipona fasciculata*, *Melipona fuscopilosa*, and *Tetragonisca angustula*). Each step has a brief explanation, as well as code and graphics.

The data preprocessing workflow used was taken from the research paper ["notame": Workflow for Non-Targeted LC–MS Metabolic Profiling](https://doi.org/10.3390/metabo10040135), which offers a wide variety of functions for performing metabolomic profile analysis.

# Before to start

The “*notame*” package accepts a feature table that can be obtained through different software, such as MZmine, MS-DIAL, and others, as input. The feature list table was extracted using the MZmine software in this case. The exported (.csv) file from MZmine was fixed to get the final feature table input according to the “*notame*” package guideline.

The raw (.csv) file modifications can be summarized by adding and renaming columns. The added columns “Column” and “Ion Mode” allow for the analysis of samples with different types of columns and different ionization polarities, respectively. Also, the cells corresponding to mass (*m/z*) and retention time (*rt*) must be renamed so the “*notame*” package can detect and process them.

# Notame workflow

The “*notame*” package and other dependency packages were installed as a first step for the analysis.

```{r echo=TRUE, message=FALSE}

# Notame package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("antonvsdata/notame", ref = "v0.3.1")

# Notame library call
library(notame)

# Dependency packages installation
install_dependencies

```

Then, a primary path and a log system were added to have a record of each process executed.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Main path
ppath <- "../Amazonia_Honey/"
# Log system
init_log(log_file = paste0(ppath, "../Result/notame_Result/HS_GCMS/HS_GCMS_log.txt"))

```

Next, the MZmine feature list table in “*notame*” format was loaded.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

data <- read_from_excel(file = "../Data/Data_to_notame/HS_GCMS_Data_to_notame.xlsx",
                        sheet = 2, corner_row = 18, corner_column = "L",
                        split_by = c("Column", "Ion Mode"))

```

Once the data was loaded, the next step was to create a MetaboSet to work with R objects from now on.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

modes <- construct_metabosets(exprs = data$exprs, 
                              pheno_data = data$pheno_data, 
                              feature_data = data$feature_data,
                              group_col = "Group")

```

## Preprocessing

The first step of the preprocessing is to change the features with a value equal to 0 to NA.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Data extraction
mode <- modes$RTX5MS_EI
# Change 0 value to NA
mode <- mark_nas(mode, value = 0)

```

Then, features with low detection rates are flagged and can be ignored or removed in subsequent analysis. The “*notame*” package uses two criteria to flag these features: the feature's presence in a percentage of QC injections and the feature's presence in a percentage within a sample group or class.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Low detection rate
mode <- flag_detection(mode, qc_limit = 7/9, group_limit = 2/3)

```

The following preprocessing step is drift correction, which is applied using smoothed cubic spline regression.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drift correction
corrected <- correct_drift(mode)
#corrected <- correct_drift(corrected)
# Flag low quality features
#corrected <- flag_quality(corrected, condition = "RSD_r < 0.3 & D_ratio_r < 0.6")

```

Then, we can visualize the data after drift correction.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
corr_bp <- plot_sample_boxplots(corrected,
                                order_by = "Group",
                                fill_by = "Species")
# PCA
corr_pca <- plot_pca(corrected,
                     center = TRUE,
                     shape = "Species",
                     color = "Species")
# Package to plots visualization in a same windows
library(patchwork)
# Plot
corr_pca + corr_bp

```

The next step is feature clustering. This step helps us reduce the number of features of the same molecule that were split due to a 70 eV electron ionization (EI) environment.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

clustered <- cluster_features(corrected, rt_window = 1/60, corr_thresh = 0.95,
                              d_thresh = 0.80)
compressed <- compress_clusters(clustered)

```

We can inspect the data using the PCA plot after the clustering algorithm execution.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
clust_bp <- plot_sample_boxplots(compressed,
                                 order_by = "Group",
                                 fill_by = "Species")
# PCA
clust_pca <- plot_pca(compressed,
                      center = TRUE,
                      shape = "Species",
                      color = "Species")
# Plot
clust_pca + clust_bp

```

## Processing

We used probabilistic quotient normalization (PQN) with missing value imputation and scaled using the autoscaling method for downstream statistical analysis.

The code below imputes the data using a chromatographic noise value.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Impute missing values using noise threshold
imputed <- impute_simple(compressed, value = 45, na_limit = 0)

```

We can inspect the data with the PCA plot after data imputation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
imp_bp <- plot_sample_boxplots(imputed,
                               order_by = "Group",
                               fill_by = "Species")
# PCA
imp_pca <- plot_pca(imputed,
                    center = TRUE,
                    shape = "Species",
                    color = "Species")
# Plot
imp_pca + imp_bp

```

After data imputation, the data were normalized by PQN.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Probabilistic quotient normalization
pqn_set <- pqn_normalization(imputed,
                             ref = c("qc", "all"),
                             method = c("median", "mean"),
                             all_features = FALSE)

```

We can inspect the data with the PCA plot after data normalization.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
pqn_bp <- plot_sample_boxplots(pqn_set,
                               order_by = "Group",
                               fill_by = "Species")
# PCA
pqn_pca <- plot_pca(pqn_set,
                    center = TRUE,
                    shape = "Species",
                    color = "Species")
# Plot
pqn_pca + pqn_bp

```

Finally, the data is ready to be exported and proceed with the statistical analysis.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

#save(compressed, file = paste0(ppath, "Result/notame_Result/HS_GCMS/Notame_HS_GC-MS_out.RData"))

```

# Principal component analysis (PCA)

Extracting the data and metadata of "*notame*" MetaboSet for principal component analysis (PCA) plot.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Extract clean data
pqn_noflag <- drop_flagged(pqn_set)
# Extracting feature height table
peak_height <- exprs(pqn_noflag)
# Extracting phenotypic data
pheno_data <- pqn_noflag@phenoData@data

```

Transposing the feature table and preparing the PCA data.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Transposing feature height table
transp_table  <- t(peak_height)
# Centering and Scaling features
ei_pca <- prcomp(transp_table, center = TRUE, scale. = TRUE)

```

## Score PCA

Plotting PCA results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 7, fig.height = 4}

# Library to left_join use
library(dplyr)
# PCA scores
scores <- ei_pca$x %>%                   # Get PC coordinates
  data.frame %>%                         # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%    # Create a new column with the sample names
  left_join(pheno_data)                  # Adding metadata
# PCA plot
pca_plot <- ggplot(scores,
       aes(PC1, PC2, shape = Species, color = Species)) +
  geom_point(size = 3) +
  scale_color_manual(values=c("#7CAE00",
                              "#F8766D",
                              "#00BFC4",
                              "#E76BF3")) +
  ggforce::geom_mark_ellipse(aes(filter = Species == "T. angustula"),
                             show.legend = FALSE, expand = unit(3, 'mm'),
                             tol = 0.01) +
  ggforce::geom_mark_ellipse(aes(filter = Species == "M. grandis"),
                             show.legend = FALSE, expand = unit(3.5, 'mm'),
                             tol = 0.01) +
  geom_text(aes(label = Site), hjust = 0, nudge_x = 0.55,
            check_overlap = TRUE, show.legend = FALSE) +
  guides(x=guide_axis(title = "PC1 (20.14 %)"),
         y=guide_axis(title = "PC2 (17.62 %)")) +
  labs(shape = 'Bees species', color= 'Bees species') +
  theme_classic() +
  theme(legend.text = element_text(face="italic")) +
  theme(legend.position = c(0.120, 0.230),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
pca_plot

```

## Loading PCA

Plotting a loading PCA result.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

loadings <- ei_pca$rotation %>%           # Extract loadings
  data.frame(Feature_ID = rownames(.))    # New column with feat name

```

Creating an artificial table with feature names and a column with compound names (identified metabolites).

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 14, fig.height = 7}

# Extracting feature identified
metab_data <- pqn_noflag[!is.na(pqn_noflag@featureData@data$Metabolite),]
# Extracting metabolite table
meta_table <- metab_data@featureData@data
# Creating a new small table of the annotated compounds
ei_compouds <- left_join(meta_table, loadings)
# Plotting results
load_pca <- ggplot(loadings, aes(PC1, PC2)) + 
  geom_point(alpha = 0.3, size = 2) +
  theme_classic() + 
  geom_point(data = ei_compouds,
             aes(shape = meta_table$IL,
                 color = meta_table$IL),
             size = 2.5) +
  labs(shape = 'Identification level',
       color = 'Identification level') +
  scale_color_manual(values = c("green",
                                "darkblue")) +
  scale_shape_manual(values = c(17, 19)) +
  ggrepel::geom_label_repel(data = ei_compouds,
                            aes(label = meta_table$Metabolite),
                            box.padding = 0.37,
                            label.padding = 0.22,
                            label.r = 0.30,
                            cex = 2.5,
                            max.overlaps = 50,
                            min.segment.length = 0) +
  guides(x=guide_axis(title = "PC1 (20.14 %)"),
         y=guide_axis(title = "PC2 (17.62 %)")) +
  theme(legend.position = c(0.070, 0.090),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
  #ggsci::scale_color_aaas()
load_pca
# Save plot
#ggsave('Result/notame_Result/HS_GCMS/load_pca.pdf', width = 14, height = 7, device='pdf', dpi="print")

```

# PERMANOVA analysis

## *T. angustula*

PERMANOVA analysis of *T. angustula* by collection site.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drop QC
pqn_noqc <- drop_qcs(pqn_noflag)
# Removal of the M. grandis group from the dataset
permanova_ta <- pqn_noqc[, pqn_noqc$Species != "M. grandis"]
pData(permanova_ta) <- droplevels(pData(permanova_ta))
# Removal of the M. fuscopilosa group from the dataset
permanova_ta <- permanova_ta[, permanova_ta$Species != "M. fuscopilosa"]
pData(permanova_ta) <- droplevels(pData(permanova_ta))
# Change to factor varibles
permanova_ta@phenoData@data$Group <- as.factor(permanova_ta@phenoData@data$Group)
# PERMANOVA between close hives of bees T. angustula site 1 and 2 (Removal of site 3)
permanova_1vs2ta <- permanova_ta[, permanova_ta$Site != "3"]
pData(permanova_1vs2ta) <- droplevels(pData(permanova_1vs2ta))
permanova_1vs2ta <- perform_permanova(permanova_1vs2ta, group = "Group", nperm = 999)
permanova_1vs2ta
# PERMANOVA between close hives of bees T. angustula site 1 and 3 (Removal of site 2)
permanova_1vs3ta <- permanova_ta[, permanova_ta$Site != "2"]
pData(permanova_1vs3ta) <- droplevels(pData(permanova_1vs3ta))
permanova_1vs3ta <- perform_permanova(permanova_1vs3ta, group = "Group", nperm = 999)
permanova_1vs3ta
# PERMANOVA between close hives of bees T. angustula site 2 and 3 (Removal of site 1)
permanova_2vs3ta <- permanova_ta[, permanova_ta$Site != "1"]
pData(permanova_2vs3ta) <- droplevels(pData(permanova_2vs3ta))
permanova_2vs3ta <- perform_permanova(permanova_2vs3ta, group = "Group", nperm = 999)
permanova_2vs3ta

```

## *M. grandis*

PERMANOVA analysis of *M. grandis* by collection site.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Removal of the T. angustula group from the dataset
permanova_mfa <- pqn_noqc[, pqn_noqc$Species != "T. angustula"]
pData(permanova_mfa) <- droplevels(pData(permanova_mfa))
# Removal of the M. fuscopilosa group from the dataset
permanova_mfa <- permanova_mfa[, permanova_mfa$Species != "M. fuscopilosa"]
pData(permanova_mfa) <- droplevels(pData(permanova_mfa))
# Change to factor varibles
permanova_mfa@phenoData@data$Group <- as.factor(permanova_mfa@phenoData@data$Group)
# PERMANOVA between close hives of bees M. grandis site 5 and 6 (Removal of site 4)
permanova_5vs6mfa <- permanova_mfa[, permanova_mfa$Site != "4"]
pData(permanova_5vs6mfa) <- droplevels(pData(permanova_5vs6mfa))
permanova_5vs6mfa <- perform_permanova(permanova_5vs6mfa, group = "Group", nperm = 999)
permanova_5vs6mfa
# PERMANOVA between close hives of bees M. grandis site 5 and 4 (Removal of site 6)
permanova_5vs4mfa <- permanova_mfa[, permanova_mfa$Site != "6"]
pData(permanova_5vs4mfa) <- droplevels(pData(permanova_5vs4mfa))
permanova_5vs4mfa <- perform_permanova(permanova_5vs4mfa, group = "Group", nperm = 999)
permanova_5vs4mfa
# PERMANOVA between close hives of bees M. grandis site 6 and 4 (Removal of site 5)
permanova_6vs4mfa <- permanova_mfa[, permanova_mfa$Site != "5"]
pData(permanova_6vs4mfa) <- droplevels(pData(permanova_6vs4mfa))
permanova_6vs4mfa <- perform_permanova(permanova_6vs4mfa, group = "Group", nperm = 999)
permanova_6vs4mfa

```

## *M. fuscopilosa*

PERMANOVA analysis of *M. fuscopilosa* by collection site.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Removal of the T. angustula group from the dataset
permanova_mfu <- pqn_noqc[, pqn_noqc$Species != "T. angustula"]
pData(permanova_mfu) <- droplevels(pData(permanova_mfu))
# Removal of the M. grandis group from the dataset
permanova_mfu <- permanova_mfu[, permanova_mfu$Species != "M. grandis"]
pData(permanova_mfu) <- droplevels(pData(permanova_mfu))
# Change to factor varibles
permanova_mfu@phenoData@data$Group <- as.factor(permanova_mfu@phenoData@data$Group)
# PERMANOVA between close hives of bees M. fuscopilosa site 8 and 9 (Removal of site 7)
permanova_8vs9mfu <- permanova_mfu[, permanova_mfu$Site != "7"]
pData(permanova_8vs9mfu) <- droplevels(pData(permanova_8vs9mfu))
permanova_8vs9mfu <- perform_permanova(permanova_8vs9mfu, group = "Group", nperm = 999)
permanova_8vs9mfu
# PERMANOVA between close hives of bees M. fuscopilosa site 8 and 7 (Removal of site 9)
permanova_8vs7mfu <- permanova_mfu[, permanova_mfu$Site != "9"]
pData(permanova_8vs7mfu) <- droplevels(pData(permanova_8vs7mfu))
permanova_8vs7mfu <- perform_permanova(permanova_8vs7mfu, group = "Group", nperm = 999)
permanova_8vs7mfu
# PERMANOVA between close hives of bees M. fuscopilosa site 9 and 7 (Removal of site 8)
permanova_9vs7mfu <- permanova_mfu[, permanova_mfu$Site != "8"]
pData(permanova_9vs7mfu) <- droplevels(pData(permanova_9vs7mfu))
permanova_9vs7mfu <- perform_permanova(permanova_9vs7mfu, group = "Group", nperm = 999)
permanova_9vs7mfu

```

# Heatmap with HCA

Only the annotated features were used in the heatmap and the hierarchical cluster analysis (HCA). The ComplexHeatmap package will be used for the heatmap and the hierarchical cluster analysis (HCA).

ComplexHeatmap package and dependency installation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# ComplexHeatmap package installation
#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
#BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)

# ColorRamp2 package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("jokergoo/colorRamp2")
library(colorRamp2)

# Cowplot package installation
#install.packages("cowplot")
library(cowplot)

# mdatools package installation
#install_github('svkucheryavski/mdatools')
library(mdatools)

# ClassyFire package installation
#remotes::install_github('aberHRML/classyfireR')
library(classyfireR)

```

The metabolites were classified using the [ClassyFireR](https://doi.org/10.1186/s13321-016-0174-y) to add these metabolite classifications to the heatmap plot.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# InChI key of the metabolites you want to classify
InChI_Keys <- c('2-Heptanone' = "CATSNJVOTSVZJV-UHFFFAOYSA-N")
# Get classification
Classification_List <- purrr::map(InChI_Keys, get_classification)
Classification_List

```

Extracting and loading of identified metabolites' abundance, and data scaling by the autoscaling method.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drop QC
hm_no_qc <- drop_qcs(pqn_noflag)
# Scaling by autoscaling method
hm_scl <- scale(t(exprs(hm_no_qc)), center = TRUE, scale = TRUE)
hm_scl <- t(hm_scl)
# Adding autoscaling data to notame MetaboSet
hm_scl_set <- hm_no_qc
exprs(hm_scl_set) <- hm_scl
# Extracting identified metabolite data
raw_hm <- hm_scl_set[!is.na(hm_scl_set@featureData@data$Metabolite),]
# Extracting feature height table
hm_height <- exprs(raw_hm)
# Extracting sample information
hm_pdata <- raw_hm@phenoData@data
# Extracting feature information
hm_fdata <- raw_hm@featureData@data

```

Row and top heatmap annotation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.width = 10, fig.height = 9}

set.seed(1540)
# Adding row and column names
hm_scl <- hm_height
rownames(hm_scl) <- hm_fdata$Metabolite
colnames(hm_scl) <- hm_pdata$Species
# Metabolite class color
cols_metclass <- c("Benzenoids" = "#800000FF",
                   "Hydrocarbons" = "#FFA319FF",
                   "Lipids and lipid-like molecules" = "#8A9045FF",
                   "Organic oxygen compounds" = "#8DD3C7",
                   "Organohalogen compounds" = "#BEBADA",
                   "Organoheterocyclic compounds" = "#FFFFB3")
# Add row anotation to HeatMap
hm_row_ann <- rowAnnotation(`Superclass` = hm_fdata$classyfireR_Superclass,
                            col = list(`Superclass` = cols_metclass),
                            show_annotation_name = T,
                            show_legend = F)
# Species color
cols_species <- c("T. angustula" = "#E76BF3",
                  "M. grandis" = "#F8766D",
                  "M. fuscopilosa" = "#7CAE00")
# Add top anotation to Heatmap
top_info_ann <- HeatmapAnnotation(`Species` = hm_pdata$Species,
                                  col = list(`Species` = cols_species),
                                  show_annotation_name = T,
                                  show_legend = F,
                                  border = T)
# Color scale
mycol <- colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))
# Heatmap matrix plotting
hm_plot <- Heatmap(hm_scl,
                   col = mycol,
                   border_gp = grid::gpar(col = "black", lty = 0.02),
                   rect_gp = grid::gpar(col = "black", lwd = 0.75),
                   clustering_distance_columns = "euclidean",
                   clustering_method_columns = "complete",
                   top_annotation = top_info_ann,
                   column_names_gp = gpar(fontface = "italic"),
                   row_names_max_width = unit(10, "cm"),
                   right_annotation = hm_row_ann,
                   show_heatmap_legend = F,
                   row_km = 3, column_km = 2,
                   row_title = c("a", "b", "c"))
hm_plot

```

Adding legends to the heatmap.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Color scale legend
lgd1 <- Legend(col_fun = mycol,
               title = "Autoscaled abundance",
               direction = "horizontal" )
# Bees species legend
lgd2 <- Legend(labels = gt_render(c("*T. angustula*",
                                    "*M. grandis*",
                                    "*M. fuscopilosa*")),
               legend_gp = gpar(fill = cols_species),
               title = "Bees species", ncol = 1)
# Metabolite class Legend
lgd3 <- Legend(labels = c(unique(hm_fdata$classyfireR_Superclass)) ,
               legend_gp = gpar(fill = cols_metclass), 
               title = "Metabolite superclass", ncol = 2)

```

Plotting the heatmap.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.width = 10, fig.height = 12}

set.seed(1540)
# Converting to ggplot
gg_heatmap <- grid.grabExpr(draw(hm_plot))
gg_heatmap <- ggpubr::as_ggplot(gg_heatmap)
# Legends
all_legends <- packLegend(lgd1, lgd2, lgd3, direction = "horizontal")
gg_legend <- grid.grabExpr(draw(all_legends))
gg_legend_fn <- ggpubr::as_ggplot(gg_legend)
# Heatmap plot
gcms_hm <- plot_grid(gg_legend_fn,
                     gg_heatmap, ncol = 1,
                     rel_heights = c(0.055, 0.880))
gcms_hm

```

# Antimicrobial activity correlation

Correlation between feature abundance and antibacterial result (zone inhibition diameter) using the PCA and Pearson's correlation approaches.

## Using the PCA

Data and metadata preparation, merging the antimicrobial data with feature abundance.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Antimicrobial activity data extraction
bacteria_data <- drop_qcs(modes$na_na)
bacteria_data <- t(exprs(bacteria_data))
# Drop QCs of metabolomics data
no_qc <- drop_qcs(pqn_noflag)
# Extracting feature height table
corr_peak <- t(exprs(no_qc))
# Merge the antimicrobial activity result with the feature list table
bacteria_peak <- cbind(bacteria_data, corr_peak)
# Extracting phenotypic data of metabolomics data
corr_pdata <- no_qc@phenoData@data

```

### Score PCA

PCA scores calculation of the antimicrobial and feature abundance merge data.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Centering and Scaling features
corr_pca <- prcomp(bacteria_peak, center = TRUE, scale. = TRUE)

```

Plotting PCA score.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 7, fig.height = 4}

# PCA scores
corr_scores <- corr_pca$x %>%            # Get PC coordinates
  data.frame %>%                         # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%    # Create a new column with the sample names
  left_join(corr_pdata)                  # Adding metadata
# PCA plot
corr_pca_plot <- ggplot(corr_scores,
                        aes(PC1, PC2, shape = Species, color = Species)) +
  scale_color_manual(values=c("#7CAE00",
                              "#F8766D",
                              "#E76BF3")) +
  scale_shape_manual(values=c(16, 17, 3)) +
  geom_point(size = 3) +
  guides(x=guide_axis(title = "PC1 (22.46 %)"),
         y=guide_axis(title = "PC2 (18.82 %)")) +
  labs(shape = 'Bees species', color= 'Bees species') +
  theme_classic() +
  theme(legend.text = element_text(face="italic")) +
  theme(legend.position = c(0.120, 0.200),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
corr_pca_plot

```

### Loading PCA

Plotting loading results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

corr_loadings <- corr_pca$rotation %>%    # Extract loadings
  data.frame(Feature_ID = rownames(.))    # New column with feat name

```

Creating an artificial table with feature names and a column with compound names (identified metabolites).

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 14, fig.height = 7}

# Extracting the antimicrobial metadata
bacteria_table <- drop_qcs(modes$na_na)@featureData@data
bacteria_table$Flag  <- as.character(bacteria_table$Flag)
# Merge the antimicrobial metadata with the metabolite table
corr_table <- merge(meta_table, bacteria_table, all = TRUE)
# Creating a new small table of the annotated compounds and antimicrobial metadata
corr_compouds <- left_join(corr_table, corr_loadings)
# Plotting results
corr_load_plot <- ggplot(corr_loadings, aes(PC1, PC2)) + 
  geom_point(alpha = 0.3, size = 2) +
  theme_classic() +
  geom_point(data = corr_compouds, aes(shape = corr_table$IL,
                                       color = corr_table$IL), size = 2.5) +
  labs(shape = 'Legend of features',
       color = 'Legend of features') +
  scale_color_manual(values = c("red",
                                "green",
                                "darkblue")) +
  scale_shape_manual(values = c(15, 17, 19)) +
  ggrepel::geom_label_repel(data = corr_compouds,
                            aes(label = corr_table$Metabolite,
                                fontface = ifelse(corr_table$IL == "Antimicrobial activity",
                                                  'italic', 'plain')),
                            box.padding = 0.37,
                            label.padding = 0.22,
                            label.r = 0.30,
                            cex = 2.5,
                            max.overlaps = 50,
                            min.segment.length = 0) +
  guides(x=guide_axis(title = "PC1 (22.46 %)"),
         y=guide_axis(title = "PC2 (18.82 %)")) +
  theme(legend.position = c(0.075, 0.127),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
  #ggsci::scale_color_aaas()
corr_load_plot

```

## Using the Pearson correlation

The Pearson correlation between antimicrobial activity and feature abundance was performed.

### Normality of the antimicrobial data

First, the antimicrobial data normality was evaluated using the Shapiro-Wilk test.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Loading the "readxl" package
library(readxl)
# Loading the antimicrobial activity data
all_bacteria <- read_excel("../Data/Antimicrobial_of_honey.xlsx", sheet = 1)
# Deleting the group column to perform Shapiro-Wilk test in batch
no_group <- all_bacteria[,-1]
# Perform the Shapiro-Wilk test
shapiro <- do.call(rbind, lapply(no_group,
                                 function(x) shapiro.test(x)[c("statistic", "p.value")]))
shapiro

```

The result showed that only the microorganisms *K. pneumonia* and *P. mirabilis* have a normal distribution (p-value > 0.05). However, when we inspected the data, we found the values in the ATA001 and RIG003 groups in the *E. coli* could be outliers; also, a possible outlier in the RYMG001 group in the *S. aureus* was found. When we delete all these possible outliers, we find that all the microorganisms (not including *S. aureus* at 25 % v/v) have a normal distribution, as seen below.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Delete outlier data in E. coli
no_outliers <- all_bacteria
no_outliers[no_outliers$Group == "ATA001",]$E_coli_75 <- NA
no_outliers[no_outliers$Group == "ATA001",]$E_coli_100 <- NA
no_outliers[no_outliers$Group == "RIG003",]$E_coli_75 <- NA
no_outliers[no_outliers$Group == "RIG003",]$E_coli_100 <- NA
# Delete outlier data in S. aureus
no_outliers[no_outliers$Group == "RYMG001",]$S_aureus_25 <- NA
no_outliers[no_outliers$Group == "RYMG001",]$S_aureus_50 <- NA
no_outliers[no_outliers$Group == "RYMG001",]$S_aureus_75 <- NA
no_outliers[no_outliers$Group == "RYMG001",]$S_aureus_100 <- NA
# Deleting the group column to perform Shapiro-Wilk test in batch
no_outliers <- no_outliers[,-1]
# Perform the Shapiro-Wilk test
shapiro_outl <- do.call(rbind, lapply(no_outliers,
                                 function(x) shapiro.test(x)[c("statistic", "p.value")]))
shapiro_outl

```

The antimicrobial activity result of honey at 75 % v/v was used in the further analysis because this concentration showed activity in most of the tested microorganisms. (*E. coli*, *K. pneumonia*, *P. mirabilis*, and *S. aureus*).

### Metabolomics data transformation

Before the Pearson correlation of feature abundance with the antimicrobial activity (zone inhibition diameter), the previously normalized data by PQN was transformed using the generalised logarithm (glog) method.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# "SummarizedExperiment" package installation
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("SummarizedExperiment")
library(SummarizedExperiment)
# Convert feature height table to SummarizedExperiment class
pmp_data <- SummarizedExperiment(assays = exprs(pqn_noflag),
                                 colData = pqn_noflag@phenoData@data)
# Package for generalized logarithmic transform
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("pmp")
library(pmp)
# Generalised logarithmic transform
glog_exprs <- glog_transformation(df = pmp_data@assays@data@listData[[1]],
                                  classes = pmp_data$QC,
                                  qc_label = "QC")
# Adding glog transformation to notame MetaboSet
glog_set <- pqn_noflag
exprs(glog_set) <- glog_exprs

```

Preparing the metabolomics data for the antimicrobial correlation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drop QCs
glog_no_qc <- drop_qcs(glog_set)
# Preparing the data for E. coli correlation (drop outlier data)
glog_E.coli <- glog_no_qc[, glog_no_qc$Group != "ATA001"]
pData(glog_E.coli) <- droplevels(pData(glog_E.coli))
glog_E.coli <- glog_E.coli[, glog_E.coli$Group != "RIG003"]
pData(glog_E.coli) <- droplevels(pData(glog_E.coli))
# Preparing the data for S. aureus (dropn outlier data)
glog_S.aureus <- glog_no_qc[, glog_no_qc$Group != "RYMG001"]
pData(glog_S.aureus) <- droplevels(pData(glog_S.aureus))
# Convert antimicrobial result to numeric
glog_E.coli$E_coli_75 <- as.numeric(glog_E.coli$E_coli_75)
glog_no_qc$K_pneumonia_75 <- as.numeric(glog_no_qc$K_pneumonia_75)
glog_no_qc$P_mirabilis_75 <- as.numeric(glog_no_qc$P_mirabilis_75)
glog_S.aureus$S_aureus_75 <- as.numeric(glog_S.aureus$S_aureus_75)
# Exporting MetaboSet data to Excel
#write_to_excel(glog_no_qc,
#               file = paste0(ppath, "../Result/notame_Result/HS_GCMS/glog_no_qc.xlsx"))

```

Perform the Pearson correlation analysis.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Feature correlation with the E. coli antimicrobial activity
E.coli_corr <- perform_correlation_tests(glog_E.coli,
                                         x = featureNames(glog_no_qc),
                                         y = c("E_coli_75"),
                                         method = "pearson")
# Feature correlation with the K. pneumonia antimicrobial activity
K.pneumonia_corr <- perform_correlation_tests(glog_no_qc,
                                              x = featureNames(glog_set),
                                              y = c("K_pneumonia_75"),
                                              method = "pearson")
# Feature correlation with the P. mirabilis antimicrobial activity
P.mirabilis_corr <- perform_correlation_tests(glog_no_qc,
                                              x = featureNames(glog_set),
                                              y = c("P_mirabilis_75"),
                                              method = "pearson")
# Feature correlation with the S. aureus antimicrobial activity
S.aureus_corr <- perform_correlation_tests(glog_S.aureus,
                                           x = featureNames(glog_no_qc),
                                           y = c("S_aureus_75"),
                                           method = "pearson")

```

### Correlation of identified metabolites

Create a matrix for the heatplot of Pearson's correlation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Correlation table for the heatmap
hm_corr_table <- data.frame(Feature_ID = E.coli_corr$X,
                            E_coli = E.coli_corr$Correlation_coefficient,
                            K_pneumonia = K.pneumonia_corr$Correlation_coefficient,
                            P_mirabilis = P.mirabilis_corr$Correlation_coefficient,
                            S_aureus = S.aureus_corr$Correlation_coefficient)
# Feature data table
feat_corr_table <- data.frame(Feature_ID = hm_fdata$Feature_ID,
                             Metabolite = hm_fdata$Metabolite)
# Adding the metabolie name to the correlation table
hm_corr_table <- left_join(hm_corr_table, feat_corr_table)
# Extracting data of the identified metabolites
hm_corr_table <- hm_corr_table[!is.na(hm_corr_table$Metabolite),]
# Adding row name
rownames(hm_corr_table) <- hm_corr_table$Metabolite
# Delete extra information
hm_corr_table <- subset(hm_corr_table, select = -c(Feature_ID, Metabolite))
# Converting DataFrame to data matrix
hm_corr_table <- data.matrix(hm_corr_table, rownames.force = NA)
# Adding column name
colnames(hm_corr_table) <- c("E. coli", "K. pneumonia", "P. mirabilis", "S. aureus")

```

Create a matrix showing the statistical significance of the correlation result.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Correlation table for the heatmap
hm_sign_table <- data.frame(Feature_ID = E.coli_corr$X,
                            E_coli = E.coli_corr$Correlation_P_FDR,
                            K_pneumonia = K.pneumonia_corr$Correlation_P_FDR,
                            P_mirabilis = P.mirabilis_corr$Correlation_P_FDR,
                            S_aureus = S.aureus_corr$Correlation_P_FDR)
# Adding the metabolie name to the correlation table
hm_sign_table <- left_join(hm_sign_table, feat_corr_table)
# Extracting data of the identified metabolites
hm_sign_table <- hm_sign_table[!is.na(hm_sign_table$Metabolite),]
# Adding row name
rownames(hm_sign_table) <- hm_sign_table$Metabolite
# Delete extra information
hm_sign_table <- subset(hm_sign_table, select = -c(Feature_ID, Metabolite))
# Converting DataFrame to data matrix
hm_sign_table <- data.matrix(hm_sign_table, rownames.force = NA)
# Adding column name
colnames(hm_sign_table) <- c("E. coli", "K. pneumonia", "P. mirabilis", "S. aureus")

```

Plot the correlation heatplot.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.width = 5, fig.height = 9}

# Add top anotation to Heatmap
ann_corr <- HeatmapAnnotation(
  foo = anno_block(gp = gpar(fill = 0, col = "white"),
                   labels = "Pearson correlation",
                   labels_gp = gpar(col = "black", fontsize = 12,
                                    fontface = "bold")))
# Color scale
col_fun <- colorRamp2(c(-1, 0, 1), c("blue", "white", "red"))
# Heatmap plot
hm_corr <- Heatmap(hm_corr_table, col = col_fun, cluster_rows = FALSE,
                   cluster_columns = FALSE, column_names_gp = gpar(fontface = "italic"),
                   right_annotation = hm_row_ann,
                   border_gp = grid::gpar(col = "black", lty = 0.02),
                   rect_gp = grid::gpar(col = "black", lwd = 0.75),
                   top_annotation = ann_corr,
                   cell_fun = function(j, i, x, y, w, h, fill) {
                     if(hm_sign_table[i, j] < 0.05) {
                       grid.text(sprintf("%.1f", hm_corr_table[i, j]), x, y,
                                 gp = gpar(fontsize = 12, fontface = "bold"))
                       } else if(hm_sign_table[i, j] > 0.05) {
                         grid.text(sprintf("%.1f", hm_corr_table[i, j]), x, y,
                                   gp = gpar(fontsize = 12))}},
                   show_heatmap_legend = F)
hm_corr

```

### Correlation of unknown metabolites

Filtering the metabolites with positive and high Pearson's correlation coefficient (r >0.5).

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Library to change character values
library('stringr')
# Correlation table for the heatmap with the best features
best_feat <- data.frame(Feature_ID = E.coli_corr$X,
                        E_coli = E.coli_corr$Correlation_coefficient,
                        E_coli_p = E.coli_corr$Correlation_P_FDR,
                        K_pneumonia = K.pneumonia_corr$Correlation_coefficient,
                        K_pneumonia_p = K.pneumonia_corr$Correlation_P_FDR,
                        P_mirabilis = P.mirabilis_corr$Correlation_coefficient,
                        P_mirabilis_p = P.mirabilis_corr$Correlation_P_FDR,
                        S_aureus = S.aureus_corr$Correlation_coefficient,
                        S_aureus_p = S.aureus_corr$Correlation_P_FDR)
# Filtering the feature with the correlation >= 0.5
best_feat <- best_feat[best_feat$E_coli >= 0.5 |
                         best_feat$K_pneumonia >= 0.5 |
                         best_feat$P_mirabilis >= 0.5 |
                         best_feat$S_aureus >= 0.5, ]
# Adding the metabolite name to the correlation table
best_feat <- left_join(best_feat, feat_corr_table)
# Extracting data of no identified metabolites
best_feat <- best_feat[!complete.cases(best_feat$Metabolite),]
# Replace the column name and ionization type "RTX5MS_EI" with unknown
best_feat$Feature_ID <- sub('RTX5MS_EI_','Unknown-', best_feat$Feature_ID)
best_feat$Feature_ID <- str_replace_all(best_feat$Feature_ID, '_', '.')

```

Create a matrix for the heatplot of Pearson's correlation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Correlation table for the heatmap
hm_corr_table1 <- data.frame(Feature_ID = best_feat$Feature_ID,
                             E_coli = best_feat$E_coli,
                             K_pneumonia = best_feat$K_pneumonia,
                             P_mirabilis = best_feat$P_mirabilis,
                             S_aureus = best_feat$S_aureus)
# Adding row name
rownames(hm_corr_table1) <- hm_corr_table1$Feature_ID
# Delete extra information
hm_corr_table1 <- subset(hm_corr_table1, select = -c(Feature_ID))
# Converting DataFrame to data matrix
hm_corr_table1 <- data.matrix(hm_corr_table1, rownames.force = NA)
# Adding column name
colnames(hm_corr_table1) <- c("E. coli", "K. pneumonia", "P. mirabilis", "S. aureus")

```

Create a matrix showing the statistical significance of the correlation result.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Correlation table for the heatmap
hm_sign_table1 <- data.frame(Feature_ID = best_feat$Feature_ID,
                             E_coli = best_feat$E_coli_p,
                             K_pneumonia = best_feat$K_pneumonia_p,
                             P_mirabilis = best_feat$P_mirabilis_p,
                             S_aureus = best_feat$S_aureus_p)
# Adding row name
rownames(hm_sign_table1) <- hm_sign_table1$Feature_ID
# Delete extra information
hm_sign_table1 <- subset(hm_sign_table1, select = -c(Feature_ID))
# Converting DataFrame to data matrix
hm_sign_table1 <- data.matrix(hm_sign_table1, rownames.force = NA)
# Adding column name
colnames(hm_sign_table1) <- c("E. coli", "K. pneumonia", "P. mirabilis", "S. aureus")

```

Plot the correlation heatplot.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.width = 5, fig.height = 6}

# Color scale legend
lgd1a <- Legend(col_fun = col_fun,
                title = "Correlation coefficient",
                direction = "horizontal" )
# Packed legends
pd = packLegend(lgd1a, direction = "horizontal")
# Heatmap plot
#png(file="Result/notame_Result/HS_GCMS/FigureA1.png", width = 5, height = 6, units = "in", res= 300)
hm_corr1 <- Heatmap(hm_corr_table1, col = col_fun, cluster_rows = FALSE,
                   cluster_columns = FALSE, column_names_gp = gpar(fontface = "italic"),
                   border_gp = grid::gpar(col = "black", lty = 0.02),
                   rect_gp = grid::gpar(col = "black", lwd = 0.75),
                   top_annotation = ann_corr,
                   cell_fun = function(j, i, x, y, w, h, fill) {
                     if(hm_sign_table1[i, j] < 0.05) {
                       grid.text(sprintf("%.1f", hm_corr_table1[i, j]), x, y,
                                 gp = gpar(fontsize = 10, fontface = "bold"))
                       } else if(hm_sign_table1[i, j] > 0.05) {
                         grid.text(sprintf("%.1f", hm_corr_table1[i, j]), x, y,
                                   gp = gpar(fontsize = 10))}},
                   show_heatmap_legend = F)
hm_corr1
draw(pd, x = unit(7.5, "cm"), y = unit(1, "cm"), just = c("left", "bottom"))
#dev.off()

```

# UpSet plot

The UpSet plot was implemented to inspect the unique or shared features of the honeys from each bee species and collection site (beehive). 

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Mean of technical replicate
upset_aver <- summary_statistics(compressed, grouping_cols = "Group")
# Dataframe with average values by the collection sites (beehive)
upset_mat <- data.frame(Feature_ID = upset_aver$Feature_ID,
                        ata001 = upset_aver$ATA001_mean,
                        er001 = upset_aver$ER001_mean,
                        fied001 = upset_aver$FIED001_mean,
                        rgb004 = upset_aver$RGB004_mean,
                        rgy001 = upset_aver$RGY001_mean,
                        rig003 = upset_aver$RIG003_mean,
                        rig005 = upset_aver$RIG005_mean,
                        rymg001 = upset_aver$RYMG001_mean,
                        ryta006 = upset_aver$RYTA006_mean)
# Adding rownames
row.names(upset_mat) <- upset_mat$Feature_ID
upset_mat$Feature_ID <- NULL
# Adding colnames
colnames(upset_mat) <- c("T. angustula ATA001",
                         "T. angustula ER001",
                         "M. fuscopilosa FIED001",
                         "M. grandis RGB004",
                         "M. fuscopilosa RGY001",
                         "M. fuscopilosa RIG003",
                         "M. grandis RIG005",
                         "M. grandis RYMG001",
                         "T. angustula RYTA006")
# Change the values by presence using one (1) or absence using 0
upset_mat[upset_mat > 0] <- 1       # presence 
upset_mat[is.na(upset_mat)] <- 0    # absence
# Make the combination matrix
comb_mat = make_comb_mat(upset_mat)
#png(file="Result/notame_Result/HS_GCMS/FigureA2_1.png", width = 7, height = 3.5, units = "in", res= 300)
upset_plot <- UpSet(comb_mat,
                    row_names_gp = gpar(fontsize = 10, fontface = "italic"),
                    top_annotation =
                      upset_top_annotation(comb_mat,
                                           gp = gpar(fill = "#F8766D",
                                                     col = "#F8766D"),
                                           add_numbers = TRUE,
                                           annotation_name_rot = 90,),
                    right_annotation =
                      upset_right_annotation(comb_mat,
                                             gp = gpar(fill = "#1C6AA8",
                                                     col = "#1C6AA8"),
                                             add_numbers = TRUE),
                    set_order = c("T. angustula ATA001",
                                  "T. angustula ER001",
                                  "T. angustula RYTA006",
                                  "M. fuscopilosa FIED001",
                                  "M. fuscopilosa RGY001",
                                  "M. fuscopilosa RIG003",
                                  "M. grandis RGB004",
                                  "M. grandis RIG005",
                                  "M. grandis RYMG001"))
#pdf(file="Result/notame_Result/HS_GCMS/FigureA2.pdf", width = 7, height = 3.5)
upset_plot
#dev.off()

```

Finish a record.

```{r}

finish_log()

```
